Relationships in the extremes and their influence on competition and coexistence

### 2021-10-29 paper/Paper.Rnw
trivial

\begin{abstract}
\noindent Understanding the coexistence of competing species is a topic of very broad 
interest in ecology. The modern coexistence
theory (MCT) of Chesson identifies several mechanisms of species coexistence,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures on species populations. However, such relationships
have typically been measured using covariance methods, which do not account for 
the possibility that environmental and competitive influences may be more correlated
when these influences are strong than when they are weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are extremely common
between ecological variables, and are important for a wide range of phenomena including 
extinction risk and the stability of ecosystem functioning. 
Here we extend the MCT to take into account the influence of ATAs. Storage
effects can be decomposed into two components, one of which quantifies the
influence of ATAs on species coexistence, and is a novel contribution of this study; and 
the other of which represents
the influence of environment-competition correlation \emph{per se}, without 
the influence of ATAs. Simulations of a simple model
%and an empirical example
illustrate that ATA influences on coexistence can be comparable in magnitude to other
mechanisms of coexistence, and that the presence or absence of ATAs can make the 
difference between species coexisting and competitive exclusion. 
\end{abstract}

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term within the decomposition provided by MCT \citep{Ellner_2019} -- see 
below for additional details.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
defined in a covariation context. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals the IGR. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Only storage effects involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is storage effects. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. To pursue goal G2, below, will quantify
the magnitude of the new ATA term in this expression, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms.

@@

%G3
To achieve goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[EC]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of values characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[EC]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[EC]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

@@

%G2 
Fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms), Fig.
\ref{fig:fig3} shows that the ATA term of (\ref{eq:full_decomp}) often, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MCT decomposition. 
Magnitudes of $\Delta_i^{[EC]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels A-D), for which the two species were equal 
competitors, storage effects (blue plus red
lines on Fig. \ref{fig:fig3}) were the only means by which coexistence could occur, and ATA effects
tended to be negative, weakening total storage effects.

>>>>

\begin{abstract}
\noindent Understanding the coexistence of competing species is a topic of very broad 
interest in ecology. The modern coexistence
theory (MCT) of Chesson identifies several mechanisms of species coexistence,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures on species populations. However, such relationships
have typically been measured using covariance methods, which do not account for 
the possibility that environmental and competitive influences may be more correlated
when these influences are strong than when they are weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATA) are extremely common
between ecological variables, and are important for a wide range of phenomena including 
extinction risk and the stability of ecosystem functioning. 
Here we extend the MCT to take into account the influence of ATAs. Storage
effects can be decomposed into two components, one of which quantifies the
influence of ATAs on species coexistence, and is a novel contribution of this study; and 
the other of which represents
the influence of environment-competition correlation \emph{per se}, 
without the influence of ATAs. Simulations of a simple model
%and an empirical example
illustrate that ATA influences on coexistence can be comparable in magnitude to other
mechanisms of coexistence, and that the presence or absence of ATAs can make the 
difference between species coexisting and competitive exclusion. 
\end{abstract}

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[E \nat C]} + \Delta_i^{[EC]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The term $\Delta_i^{[EC]}$ is the component of storage effects due to $EC$ correlation \emph{per se}, remaining after 
having removed the effects of ATAs.
The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term within the decomposition provided by MCT \citep{Ellner_2019} -- see 
below for additional details.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, whereas ATAs are 
defined in a covariation context. Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals the IGR. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Only storage effects involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is storage effects. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. To pursue goal G2, below, will quantify
the magnitude of the new ATA term in this expression, $\Delta_i^{[E \nat C]}$, relative to the magnitudes of the other terms.

@@

%G3
To achieve goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of values characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

@@

%G2 
Fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms), Fig.
\ref{fig:fig3} shows that the ATA term of (\ref{eq:full_decomp}) often, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MCT decomposition. 
Magnitudes of $\Delta_i^{[E \nat C]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels A-D), for which the two species were equal 
competitors, storage effects (blue plus red
lines on Fig. \ref{fig:fig3}) were the only means by which coexistence could occur, and ATA effects
tended to be negative, weakening total storage effects.
